Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-07 manuscript.tex
edits up to methods, removal of NC from study

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection is expected to remove linked neutral variation and reduce the $N_{e}$ of the Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the underlying causes of low Y-linked nucleotide diversity. Here, we investigate the factors affecting X- and Y-linked neutral polymorphsm during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae), a dioecious annual with recently evolved sex chromosomes. Strikingly, we find that neutral polymorphsm for genes on the Y is on average \textasciitilde 2.1\% of the value for their homologues on the X, corrsponding to a chromosome-wide reduction of \textasciitilde 93\% compared to the neutral expectation. We demonstrate that the magnitude of this diversity loss is inconsistent with a reduced male $N_{e}$ caused by neutral processes including female-biased sex ratios and high variance in male reproductive success. Rather, using forward population genetic simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity can be explained by purifying selection removing linked neutral variation, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that Y chromosome degeneration in the early stages occurs through strong selective interference among many linked sites rather than primarily through Y-inactivation followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plant and animal kingdoms \citep{westergaard1958,ohno1967,bull1983,charlesworth1991}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species this genetic degeneration has led to the evolution dosage compensation \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}, and reduced genetic variation on Y chromosomes \citep{qiu2010nucleotide, ellegren2011,bachtrog2013NRG}. The independent evolution of these phenomena in a broad range of species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a difficult challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.

@@

In species with female-biased reproductive sex ratios or extensive male-male competiton, a high variance in male reproductive success could also reduce the $N_{e}$ of the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting the need to distinguish the effects of variance in reproductive success from the effects selective interference just mentioned. Because high variance in male reproductive success is expected to reduce both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can be obtained by examining levels of neutral variability on X and Y chromosomes relative to values for the autosomes. In particular, high variance in male reproductive success predicts a reduction in the Y/A diversity ratio, but an increase in X/A. Based on this comparison, studies of the X/A ratio in humans have suggested a historical excess of breeding females over males resulting in an inflated X/A ratio (though there is considerable variation in estimates of X/A in humans, and other demographic scenarios have also been suggested \citep{hammer2010,bustamante2009}).
%as modeled by \citep{kimura1964number} and \citep{nomura2002effective}.

%In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which causes a reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. The major types of HRI include (i) selection on strongly beneficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} (ii) selection against strongly deleterious mutations) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}, and (iii) intereference between weakly selected sites (weak selection Hill-Robertson interference). In all of these processes, linkage  disequilibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection.

%Consistent with an important role for Hill-Robertson interference in driving down diversity on the Y chromosome, studies examining Y chromosome variability in mammals \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution, with evidence for linked selection playing a key role.

%discussion point?:
%Moreover, because recombination suppression was relatively recent on plant Y chromosomes, it is an open question whether weak-HRI has had enough time to have a large effect. Although selective interference (indlucing BGS and Selective Sweeps) is expected to have the greatest effects during the early stages of Y-chromosome evolution \citep{bachtrog2008temporal} the R. hastatulus Y chromosome has undergone significant degeneration and gene loss over a short period of time, suggesting that it may behave more similarly to more ancient Y chromosomes.

%Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes. On the other hand, if gene silencing and Y inactivation occur very early during the process of Y chromosome degeneration, it is possible that the strength of linked selection is minimized rapidly even on relatively young Y chromosomes, and that much degeneration may occur neutrally following the silencing of Y-linked genes (cite backtrog NRG review).

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with highly heteromorphic X and Y chromosomes that are estimated to have evolved  approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making sex chromosomes in this species over 100 million years younger than the well-studied mammalian sex chromosomes \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the unique occurrence of an intraspecific sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct 'chromosomal races' \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species (the North Carolina race) is thought to have originated through an X-autosome fusion, with the XY system (the Texas race) maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, a recent study revealed that both the ancestral and neo-Y chromosomes have undergone significant gene loss and functional deterioration \citep{hough2014}.

Of particular relevance to the present study, \textit{R. hastatulus} populations have also been found to consistently exhibit female-biased sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Using genome-wide estimates of neutral polymorphism together with forward population genetic simulations, we test whether levels of genetic diversity on sex chromosomes and autosomes in this species are consistent with neutral evolutionary processes, including the joint effects of sex ratio bias and variance in reproductive success, or whether the removal of linked neutral variation caused by purifying selection is also playing a significant role.

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection is expected to remove linked neutral variation and reduce the $N_{e}$ of Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y diversity. Here, we investigate the factors affecting X- and Y-linked neutral polymorphsm during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae), a dioecious annual with recently evolved sex chromosomes. Strikingly, we find that neutral diversity for genes on the Y is on average \textasciitilde 2.1\% of the value for their homologues on the X, corrsponding to a chromosome-wide reduction of \textasciitilde 93\% compared to the neutral expectation. We demonstrate that the magnitude of this diversity loss is inconsistent with a reduced male $N_{e}$ caused by neutral processes including female-biased sex ratios and high variance in male reproductive success. Rather, using forward population genetic simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity can be explained by purifying selection removing linked neutral variation, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that Y chromosome degeneration in the early stages occurs through strong selective interference among many linked sites rather than primarily through Y-inactivation followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plant and animal kingdoms \citep{westergaard1958,ohno1967,bull1983,charlesworth1991}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation of deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species, this genetic degeneration of the Y has led to the evolution dosage compensation of the X \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}. The independent evolution of these phenomena across a broad range of distantly-related species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a longstanding challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.
% and reduced genetic variation on Y chromosomes \citep{qiu2010nucleotide, ellegren2011,bachtrog2013NRG}.

@@

In species with female-biased reproductive sex ratios or extensive male-male competiton, a high variance in male reproductive success can also reduce the $N_{e}$ of genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects of linked selection on the Y need to be distinguished from neutral models that include sex ratio variation as well as male-female differences in reproductive success. High variance in male reproductive success, for example, is predicted to reduce both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, and therefore evidence for this effect can be obtained by comparing levels of neutral variability on X and Y chromosomes relative to the autosomes. In particular, high variance in male reproductive success predicts a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans have found examples of historical sex-differences in dispersal during human migration \citep{wilkins2006unraveling}, and the inflated X/A ratio has been suggested to reflect a historical excess of breeding females over males \citep{hammer2008sex}, though there is considerable variation in X/A estimates in human and other demographic scenarios have also been suggested \citep{bustamante2009,hammer2010}.

Despite widespread interest in determining the effects of selection patterns of sex-linked variability diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about how linked selection has affected patterns nucleotide polymorphism. In humans and Drosophila \citep{mcallister1999,bachtrog2000}, where large population genomic data sets are readily available, studies examining Y chromosome variability have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014}. However, it remains unclear whether such inferences about the effects of linked selection on the Y should also apply to systems in which sex chromosomes evolved \textit{do novo} from autosomes much more recently. The effects of diversity loss due to selection might be expeted to be lower in such systems because of a shorter history of recombination suppression, but contrary to this, simulations reveal that background selection and selective sweeps both have strongest effects during the early stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}. Indeed, studies of Y-degeneration in the young plant sex chromosomes \textit{R. hastatulus} and \textit{Silene latifolia} have both revealved signs of rapid Y chromosome degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. In these cases, if the early stages of Y-chromosome evolution involved widespread Y-linked gene-inactivation (silencing) \citep{orr1998adaptive}, then it is possible that the strength of linked selection would then be minimized such that gene loss was mostly driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG} rather than selective interference due to widespread linkage disequilibrium.

%The strength of this effect is dependent on the rate of recombination between a selected locus and a neighbouring region; essentially, the effect becomes more pronounced with less recombination because of linkage disequilibrium between adjacent loci.

%On the other hand, if gene silencing and Y inactivation occur very early during the process of Y chromosome degeneration, it is possible that the strength of linked selection is minimized rapidly even on relatively young Y chromosomes, and that much degeneration may occur neutrally following the silencing of Y-linked genes (cite backtrog NRG review).

%because recombination suppression was relatively recent on plant Y chromosomes, it is an open question whether weak-HRI has had enough time to have a large effect. Although selective interference (indlucing BGS and Selective Sweeps) is expected to have the greatest effects during the early stages of Y-chromosome evolution \citep{bachtrog2008temporal} the R. hastatulus Y chromosome has undergone significant degeneration and gene loss over a short period of time, suggesting that it may behave more similarly to more ancient Y chromosomes.

%Background selection (purifying selection against deleter- ious alleles) [27] and selective sweeps (positive selection for beneficial variants) [28] both have the effect of reducing nucleotide diversity at linked sites.
%as modeled by \citep{kimura1964number} and \citep{nomura2002effective}.

%evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which causes a reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. The major types of HRI include (i) selection on strongly beneficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} (ii) selection against strongly deleterious mutations) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}, and (iii) intereference between weakly selected sites (weak selection Hill-Robertson interference). In all of these processes, linkage  disequilibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection.

%Consistent with an important role for Hill-Robertson interference in driving down diversity on the Y chromosome, s

%discussion point?:
%Moreover,

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with highly heteromorphic X and Y chromosomes that are estimated to have evolved  approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making sex chromosomes in this species over 100 million years younger than the well-studied mammalian sex chromosomes \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the unique occurrence of an intraspecific sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct 'chromosomal races' \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species (the North Carolina race) is thought to have originated through an X-autosome fusion, with the XY system (the Texas race) maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, a recent study revealed that both the ancestral and neo-Y chromosomes have undergone significant gene loss and functional deterioration \citep{hough2014}.  In order to simplify our comparison of polymorphism levels within the genome, we focus on the ancestral Y chromosome that occurs in both sex chromosome races.

Of particular relevance to the present study, \textit{R. hastatulus} populations have been found to consistently exhibit female-biased reproductive sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Female-biased sex ratios in dioecious plants with sex chromosomes are not uncommon (refs), and their occurrence in \textit{Rumex} provides an excellent opportunity to study both the demographic and selective effects contributing to sex-linked variability. Here, using genome-wide estimates of neutral polymorphism together with forward population genetic simulations, we test whether levels of genetic diversity on sex chromosomes and autosomes in this species are consistent with neutral evolutionary processes, including the joint effects of sex ratio bias and high variance in reproductive success, or whether the removal of linked neutral variation caused by selection on the Y has led to a loss of diversity greater than expected under neutrality.
